---
title: "Codebook"
author: "Coleen Smith"
date: "12 April 2019"
output:
    github_document:
      toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Description  

This assignment will create one R script called run_analysis.R that does the following.  

    1. Merges the training and the test sets to create one data set.
    2. Extracts only the measurements on the mean and standard deviation for each measurement.
    3. Uses descriptive activity names to name the activities in the data set.
    4. Appropriately labels the data set with descriptive variable names.
    5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject. 

###Notes on the original (raw) data 

The data for this project came from the [Human Activity Recognition Using Smartphones Data Set](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) that was "built from the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors." From: Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. A Public Domain Dataset for Human Activity Recognition Using Smartphones. 21th European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

[Full description](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)  [Link to data](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip)

##Creating the tidy data set (final_tidyframe)

For this assignment, I merged data files from the original UCI Data Set containing, information about subject, activity and a set of 561 features measurements. The final product was a tidy data set (final_tidyframe) that represeneted the average of each variable for each combination of activity and subject.

For a complete description of the steps to complete the assignment, please review [Readme](https://github.com/Likes2Weave/datasciencecoursera/blob/master/Course3/Readme.md) available in the GitHub repository. 

``` {r read tidy file}
    final_tidyframe <- read.table("final_tidyframe.csv", header = TRUE, sep = ",")
```

##Description of the variables in the tiny_data.txt file

Dimensions: 180 observations of 81 variables.
Variables: 
``` {r variable list}
    ## show names
    names(final_tidyframe)
```

In his paper, "Tidy Data," Wickham discusses the order in which variables should appear in a data set:

>"Fixed variables describe the experimental design and are known in advance... Measured variables are what we actually measure in the study. Fixed variables should come first, followed by measured variables, each ordered so that related variables are contiguous."

By this standard, subject and activityName would be fixed variables and should appear contiguously and first.  Measured variables would be the items described by the feature measurements included in the original data set. 

### Fixed Variables

#### subject
Class: Integer
Values: Range from 1 -30
Source: UCI HAR Data Set subject_test.txt and subject_train.txt

#### activityName
Class: Factor with 6 levels
Values: LAYING, SITTING, STANDING, WALKING, WALKING_DOWNSTAIRS, WALKING_UPSTAIRS
Source: UCI HAR Data Set activity_labels.txt
Notes: Assigned to observations based on the activity number from Y_test.txt and Y_train.txt mapped to the activity number and name pair in activity_labels.txt

### Measured Variables
The remaining variables in column 3-81 are computed values.
Class: Numeric
Values: calculated average of the Source mean and standard deviation measurements
Source: UCI HAR Data Set X_test.txt and X_train.text
Notes: For a detailed summary of the original data, review the [ReadMe.txt](https://github.com/Likes2Weave/datasciencecoursera/blob/master/Course3/UCI_HAR_Dataset/features_info.txt) included in the original data set.

``` {r summary}
    summary(final_tidyframe[3:81])
```

